---
title: "pestr Workflow"
author: "Michal Jan Czyz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pestr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library('kableExtra')
library('dplyr')
```

# Overview

**pestr** Package is a set of functions and wrappers that allow painless and quick data retrival on pests and their hosts from [EPPO Data Services](https://data.eppo.int/). First of all it allow extraction of *scientific names* of organisms (and viruses), as well as *synonyms* and *common names* from SQLite database. Second, there are three functions in the package that use REST API to extract data on *hosts*, *categorization* and *taxonomy*. Lastly there is a function that downloads \*.csv files containing information on organisms (and viruses) distribution. Besides functions that deal with extracting data and putting them into tables in *human friendly* format, we support users with functions that allow to download (**this function works correctly only under LINUX**) and connect to file with SQLite database.

# Example workflow

## Setting up `token` and connecting to **SQLite db**
In order to start working with **pestr** package, you should register yourself (*free of charge*) for EPPO Data Services. Than you should run `create_eppo_token()` which will store your credentials in variable used by functions that connect to REST API.

```{r eval = F}
create_eppo_token('<<your_EPPO_token')
```

Next if you are on **Linux** operating system you can run `eppo_database_download()` function that will automaticaly download archive with the **SQLite** file and extract it into your working directory (you can overide default behaviour).

```{r eval = F}
eppo_database_download()
```

When working on **Windows** you can manually [download](https://data.eppo.int/files/sqlite.zip) and extract the archive.

Last step of setup is to connect to database file, which can be easily done with `eppo_database_connect` function.

```{r eval = F}
eppo_SQLite <- eppo_database_connect()
```

After this three short steps you are ready to go.

## Extracting tables with scientific, common and synonym names



```{r eval = F}


# Create vector of names that you are looking for
pest <- 'Helicoverpa zea'
# Connect to SQLite db


# Get pest names, in result you will have list containing 4 tables:
# df with names that are present in EPPO, df with names that are not present
# df with preferred names and eppo codes, df with all associated names to 
# eppocode from third df. It is necessary to run it before other 'tabletools'
# functions since it extracts eppocodes that are used later by other functions
pest_names <- eppo_names_tables(pest, eppo_SQLite)

# Get pest categorization - as result you will get list with two elements:
# first is a list of tables for each eppocode one categorisation table, second
# element is single df with categorization for each eppocode condensed to single
# cell.
pests_cat <- eppo_tabletools_cat(pest_names, eppo_token)
cat_table <-  pests_cat[[1]][[1]] %>% select(c('nomcontinent','country','qlistlabel','yr_add'))

# Get pest hosts as a result you get two tables: firs is long table with all data 
# for all pests combined. In second hosts are combined into single cell for each
# eppocode
pests_hosts <- eppo_tabletools_hosts(pest_names, eppo_token)

hosts_tab <- pests_hosts[[1]] %>% rename(Comments = labelclass) %>% mutate(Presence_PRA_area = 'Yes/no',Reference = '@eppo2018') %>% select(full_name, Presence_PRA_area, Comments, Reference)

# Get taxonomy. Also list with two elements. Fist is a list of taxonomy tables
# for each pest; second is table with 'main category' of each eppocode 
pests_taxo <- eppo_tabletools_taxo(pest_names, eppo_token)

# Get distributions of hosts. As a result a two element list. First one contains
# df of distribution for each eppocode, second contains single cell of distribution
# for each eppocode.
pest_distri <- eppo_tabletools_distri(pest_names)

# Whole condensed table in one shot:
eppo_fulltable <- eppo_table_full(pests, eppo_SQLite, eppo_token)
eppo_fulltable2 <- eppo_table_full(pests2, eppo_SQLite, eppo_token)

kable(pests_cat[[1]] %>%
        bind_rows(.id = 'eppocode')) %>%
  kable_styling(full_width = T, bootstrap_options = c('striped', 'condensed', 'bordered')) %>%
  row_spec(row = 0, bold = T, background = '#F5F6FA') %>%
  collapse_rows(columns = 1:2, valign = "middle")
kable(pests_hosts[[2]]) %>%
  kable_styling(full_width = T, bootstrap_options = c('striped', 'bordered')) %>%
  row_spec(row = 0, bold = T, background = '#F5F6FA')
eppo_fulltable2 %>%
  slice(-1L) %>%
  select(-'codeid', - 'eppocode') %>%
  kable() %>%
  kable_styling(full_width = T, bootstrap_options = c('striped', 'bordered')) %>%
  row_spec(row = 0, bold = T, background = '#F5F6FA')



## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.


Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
