---
title: "Example workflow of checking taxons affecting Abies alba"
output:
  rmarkdown::html_vignette:
    css: alert_style.css
vignette: >
  %\VignetteIndexEntry{Example workflow of checking taxons affecting Abies alba}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.show = "hold",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library("pestr")
library("dplyr")
library("stringr")
```

# Overview

Assume that you are assesing pest threats to *Abies alba* -- **European silver fir**. It would be wise to go to [EPPO Global Database](https://gd.eppo.int), type *Abies alba* in search field and check all the infromations you need, copy paste them into spreadsheet and make some analyses. On the other hand, you can just use `pestr` package and do everything easily from `R` directly saving time on whole *copy-pasting* procedures. To make this example easy lets assume that your goal is to check number of taxons and number of particular types (*major*, *minor*, etc.) of pests. Therefore steps you would need to make this analysis in `R` are presented below.

<div class="alert alert-warning">
I assume that you already have *eppocodes.sqlite* database in your *working directory*. If not you should download it manually from [EPPO Data Services](https://data.eppo.int) or using `eppo_database_download()` function. I also assume that you have basic knowledge of using `dplyr` and `stringr` package and piping `%>%` operator.
</div>

## Step I -- get valid *eppocode* for *Abies alba*

First you need valid *eppocode* of *Abies alba*. To do this you will need to use `eppo_names_table()` function and save your results to a variable like in the code example below:

```{r eval = FALSE, echo = TRUE}

abies_alba <- eppo_names_tables("Abies alba")
```

This variable will be our input for next step.

## Step II -- query **EPPO Data Services**

We can now look for all the existing pests of *Abies alba* that exists in [EPPO Data Services](https://data.eppo.int). To do this we need to use `eppo_tabeltools_pests()` function. It takes two *arguments* -- first is result of `eppo_names_tables()` function (which are stored in `abies_alba` variable) and the second one is token -- which allows you to connect to **EPPO API**. You get it by registering to [EPPO Data Services](https://data.eppo.int) which is free of charge. You need to paste your token to `create_eppo_token()` function - it will create variable in your `R Global Environment` that will be used among all `pestr` functions that connect to **EPPO API**. You need to do it just once per session. Please refer to code below:

```{r eval = FALSE, echo = TRUE}
### Firsr create eppo_token variable
eppo_create_token("<place your eppo token here>")

### For second argument (token) please use eppo_token in eppo_tabletools_*
### function family
abies_alba_pests <- eppo_tabletools_pests(abies_alba, eppo_token)
```

```{r eval = TRUE, echo = FALSE}
abies_alba_pests <- readRDS("abies_alba_pest.RDS")
```

Looking at structure of `abies_alba_pests` variable we see that it is a *list* containing 2 *data frames*.

```{r eval = TRUE, echo = TRUE}
str(abies_alba_pests)
```

The *long_table* element contains what we actually need for our analyses: *pest_eppocode* and *labelclass* columns. Lets start with latter.

## Step 3 -- number of pests type

Pest types are stored in *labelclass* column. We will use few very basic `R` and `dplyr` package commands to check numbers we are interested in.

```{r eval = TRUE, echo = TRUE}
### First select colum labelclass from long_table element,
### and use table to check frequencies
abies_alba_pests$long_table %>%
  dplyr::select(labelclass) %>%
  table

### Than we can create very simple barplot to visualise number of
### species in particular type of pest
abies_alba_pests$long_table %>%
  dplyr::select(labelclass) %>%
  table %>%
  barplot(xlab = "Type of pest", ylab = "Number of species",
           col = "#AF0011", ylim = c(0, 30))


```

## Step 4 -- Number of taxons

Obtaining number of pests taxons is slightly more complicated. In the current version functions that communicate with **REST API** are using output of `eppo_names_table()` function and not directly *eppocodes*. One of the reasons is that scientific names often changes as well as taxonomical classification. Thus safer method is to query most updated database with common or scientific names and get all the valid eppocodes from database. It would be strightforward to just take *fullnames* column and parse it an argument to `eppo_names_table()` function. However as we see some names in *pests* has suffix (eg. *Melampsora medusae (as Abies)*). So first we need to fix this. We will do this with *regular expressions* or *regex* as you will. You can see that the part of string we do not need starts with *space* than there is *(* some *characters and spaces*, and it ends with *)*. Using `stringr` package we will create *regex* that will be subsituted with *nothing*. Our regex expression look below:

> " \\\\([:print:]*\\\\)" <p><p>
> which translates to: <p><p>
> "space left-parenthesis any number of letters, numbers, punctuation and whitespaces right-parenthesis"

For detailed description of *regex* in `R` please go to [article on stringr package](https://stringr.tidyverse.org/articles/regular-expressions.html).

```{r eval = TRUE, echo = TRUE}
### Extract pest names and get rid of additional information in parentheses
pests_names <- abies_alba_pests$long_table %>%
  .$fullname %>%
  stringr::str_replace_all(" \\([:print:]*\\)", "")
### We also extract eppocodes to use later
pests_eppocodes <- abies_alba_pests$long_table %>%
  .$pests_eppocode
```

We now have two important variables `pests_names` and `pests_eppocodes` which will help us in quering database and **REST API**. First we query the databes with our `pest_names`:

```{r eval = FALSE, echo = TRUE}
pests_names_tables <- eppo_names_tables(pests_names)
```

Than we use our `pest_names_tables` variable as input for `eppo_tabletools_taxo()` function which will return us taxonomy.

```{r eval = FALSE, echo = TRUE}
pests_taxonomy <- eppo_tabletools_taxo(pests_names_tables, eppo_token)
```

```{r eval = TRUE, echo = FALSE}
pests_taxonomy <- readRDS("pests_taxonomy.RDS")
```

This time we can take a shortcut and use second element of list `compact_table`. Since matching names in database is not exact, it means that we sometimes obtain "more" organisms -- like main species and subspecies that have different *eppocodes*. That is why we previously saved pests *eppocodes* so now we can filter results, check the numbers using `table()` function and plot results with simple `barplot`.

```{r eval = TRUE, echo = TRUE}
pests_taxonomy$compact_table %>%
  dplyr::filter(eppocode %in% pests_eppocodes) %>%
  dplyr::select(taxonomy) %>%
  table()

pests_taxonomy$compact_table %>%
  dplyr::filter(eppocode %in% pests_eppocodes) %>%
  dplyr::select(taxonomy) %>%
  table() %>%
   barplot(xlab = "Classification of pest", ylab = "Number of species",
           col = "#AF0011", ylim = c(0, 30))
```

Now nowing basics of how functions interacts, you can play with your own workflows and analyses.
