---
title: "pestr Workflow"
author: "Michal Jan Czyz"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pestr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library('dplyr')
```

# Overview

**pestr** Package is a set of functions and wrappers that allow painless and quick data retrival on pests and their hosts from [EPPO Data Services](https://data.eppo.int/). First of all it allow extraction of *scientific names* of organisms (and viruses), as well as *synonyms* and *common names* from SQLite database. Second, there are three functions in the package that use REST API to extract data on *hosts*, *categorization* and *taxonomy*. Lastly there is a function that downloads \*.csv files containing information on organisms (and viruses) distribution. Besides functions that deal with extracting data and putting them into tables in *human friendly* format, we support users with functions that allow to download (**this function works correctly only under LINUX**) and connect to file with SQLite database.

# Example workflow

## Setting up `token` and connecting to **SQLite db**
In order to start working with **pestr** package, you should register yourself (*free of charge*) for EPPO Data Services. Than you should run `create_eppo_token()` which will store your credentials in variable used by functions that connect to REST API.

```{r eval = F}
create_eppo_token('<<your_EPPO_token')
```

Next if you are on **Linux** operating system you can run `eppo_database_download()` function that will automaticaly download archive with the **SQLite** file and extract it into your working directory (you can overide default behaviour).

```{r eval = F}
eppo_database_download()
```

When working on **Windows** you can manually [download](https://data.eppo.int/files/sqlite.zip) and extract the archive.

Last step of setup is to connect to database file, which can be easily done with `eppo_database_connect()` function.

```{r eval = F}
eppo_SQLite <- eppo_database_connect()
```

After this three short steps you are ready to go.

## Extracting tables with scientific, common and synonym names

Currently searching for pest names supports scientific names, synonyms and common names. By default search will work with
partial names -- e.g. when you query for *Cydia packardi* you get infomation related to this species, while when you query for *Cydia* you get information on whole genus. Likewise when you search for *Droso* you will get information on all species that contain *Droso* in their names. Moreover you can parse whole vector of terms in one shot, e.g. `c('Xylella', 'Cydia packardi', 'Droso')`.

```{r eval = F}
# Create vector of names that you are looking for
pests_query <- c('Helicoverpa zea', 'Cydia')
```

Than you should start with quering for names and assigning your results to a variable. This variable will contain `eppocodes` that are used by other functions to extract data from EPPO REST API. `eppo_names_tables()` takes two arguments: first is a vector of names to query the database, second is variable with connection to **SQLite** database.

```{r eval = F}
pest_names <- eppo_names_tables(pests_query, eppo_SQLite)
```

As a result you will get  list containing 4 tables:
* df with names that are present in EPPO,
* df with names that are not present in database
* df with preferred names and eppo codes,
* df with all associated names to eppocode from third df. 

**REMEMBER:** It is necessary to run it before other 'tabletools' functions since it extracts eppocodes that are used later by other functions.

## EPPO *tabletools* functions to extract categorization, hosts, taxonomy and distribution

This functions works separately from each other, thus there is no need to use all of them as well as ther is no need to use them in any particular order. Functions for categorization, hosts and taxonomy takes two arguments: a varibale containing result of `eppo_names_tables()` and variable with EPPO token. 

### categorization
As result `eppo_tabletools_cat()` you will get list with two elements: first is a list of tables -- for each eppocode one categorisation table, second element is a single df with categorization for each eppocode condensed to single cell.

```{r eval = F}
pests_cat <- eppo_tabletools_cat(pest_names, eppo_token)
```

### hosts
`eppo_tabletools_hosts()` as a result gives two tables: first is long table with all the data, for all pests combined. In second hosts are combined into single cell for each eppocode.

```{r eval = F}
pests_hosts <- eppo_tabletools_hosts(pest_names, eppo_token)
```

### taxonomy

In the case of `eppo_tabletools_taxo()` you will also get a list with two elements. Fist is a list of taxonomy tables for each pest; second is table with 'main category' of each eppocode.

```{r eval = F}
pests_taxo <- eppo_tabletools_taxo(pest_names, eppo_token)
```

### distribution
`eppo_tabletools_distri()` does not connect to REST API, but it downloads \*.csv files from EPPO. Thus it takes only one argument (since it does not need the EPPO token) -- a varibale containing result of `eppo_names_tables`. As a result a two element list. First one contains df of distribution for each eppocode, second contains single cell of distribution for each eppocode. The distribution data is grouped within cell by continent.

```{r eval = F}
pest_distri <- eppo_tabletools_distri(pest_names)
```

## Whole condensed table in one shot:

If you want to make one table with all the infromation: names, categorization, hosts, distirbution and taxonomy -- condensed to one cell *per* pest, please use `eppo_table_full()` function that takes arguments: vector of names, **SQLite** connection and EPPO token. 

```{r eval = F}
eppo_fulltable <- eppo_table_full(pest_names, eppo_SQLite, eppo_token)
```
